Packaging research artefacts with RO-Crate

### 2022-01-01 latex/ro-crate.tex
avoid schema.org as hyperlinks

\begin{review}{solicited}
\reviewer{\fnms{Angelo A.} \snm{Salatino}\ead
[label=os1,orcid]{0000-0002-4763-3943}\printead[presep={\ }]{os1}}
\reviewer{\fnms{Alessia} \snm{Bardi}\ead
[label=os2,orcid]{0000-0002-1112-1292}\printead[presep={\ }]{os2}}
\reviewer{\snm{one anonymous reviewer}}
\end{review}
%
%\dedicated{}
%
% HISTORY:
\received{\sday{13} \smonth{8} \syear{2021}}
\accepted{\sday{23} \smonth{9} \syear{2021}}
%
\begin{abstract}
%\parttitle{}
An increasing number of researchers support reproducibility by
including pointers to and
descriptions of datasets, software and methods in their publications.
However, scientific
articles may be ambiguous, incomplete and difficult to process by
automated systems. In
this paper we introduce RO-Crate, an open, community-driven, and
lightweight approach
to packaging research artefacts along with their metadata in a machine
readable manner.
RO-Crate is based on \surl{Schema.org} annotations in JSON-LD, aiming to
establish best
practices to formally describe metadata in an accessible and practical
way for their
use in a wide variety of situations.

@@

For our development methodology we followed the mantra of working
closely with a small group to really get a deep understanding of
requirements and ensure rapid feedback loops. We created a pool of
early adopter projects from a range of disciplines and groups,
primarily addressing developers of platforms. Thus the base level for
simplicity was \textbf{developer friendliness}.

@@

Addressing the simplicity of understanding and engaging with RO-Crate
by data practitioners is through the platforms, for example with
interactive tools (Section~\ref{sec:tooling}) like
Describo\footnote{\url{https://uts-eresearch.github.io/describo/}} \cite{describo} and
Jupyter notebooks \cite{doi:10.3233/978-1-61499-649-1-87}, and by close
discussions with domain scientists on how to appropriately capture what
they determine to be relevant metadata. This ultimately requires a new
type of awareness and learning material separate from developer
specifications, focusing on the simplicity of extensibility to serve
the user needs, along with user-driven development of new RO-Crate
Profiles specific for their needs (Section~\ref{sec:inuse}).

@@

\begin{figure}%[t!]
\includegraphics{210053f02}
\caption{Simplified UML class diagram of RO-Crate. The \emph
{RO-Crate Metadata File} conforms to a version of the specification;
and contains a JSON-LD graph \cite{sporny_2014} that describes the
entities that make up the RO-Crate. The \emph{RO-Crate Root Data
Entity} represent the Research Object as a dataset. The RO-Crate
aggregates \emph{data entities} (\texttt{hasPart}) which are further
described using \emph{contextual entities} (which may include
aggregated and non-aggregated data entities). Multiple types and
relations from \surl{Schema.org} allow annotations to be more specific,
including figures, nested datasets, computational workflows, people,
organisations, instruments and places. Contextual entities not
otherwise cross-referenced from other entities' properties (\emph
{describes}) can be grouped under the root entity (\texttt{mentions}).}
\label{fig:uml}
\end{figure}

@@

As such, the RO-Crate 1.1\footnote{\url{https://w3id.org/ro/crate/1.1}}
specification \cite{doi:10.5281/zenodo.4541002} can be seen as an
opinionated and example-driven guide to writing
\surl{Schema.org}\footnote{\url{https://schema.org/}} \cite{doi:10.1145/2857274.2857276}
metadata as JSON-LD \cite{sporny_2014} (see Section~\ref{sec:implementation}), which leaves it open for implementers to include
additional metadata using other \surl{Schema.org} types and properties, or
even additional Linked Data vocabularies/ontologies or their own ad-hoc terms.

@@

The RO-Crate conceptual model has been realised using JSON-LD and
\surl{Schema.org} in a prescriptive form as discussed in Section~\ref{sec:conceptual}. These technical choices were made to cater for
simplicity from a developer perspective (as introduced in Section~\ref{sec:methodology}).

@@

In\footnotetext{Recommended properties for types shown in
Listing \ref{lis1} also include
\texttt{affiliation}, \texttt{citation}, \texttt{contactPoint},
\texttt{description},
\texttt{encodingFormat}, \texttt{funder}, \texttt{geo}, \texttt{identifier},
\texttt{keywords},
\texttt{publisher}; these properties and corresponding contextual
entities are
excluded here for brevity. See complete example
\url
{https://www.researchobject.org/2021-packaging-research-artefacts-with-ro-crate/listing1/}.} this flattened profile of JSON-LD, each \texttt{{entity}} are directly
under \texttt{@graph} and represents the RDF triples with a common \textit{subject}
(\texttt{@id}), mapped \textit{properties} like \texttt{hasPart}, and \textit{objects} -- as
either literal \texttt{"string"} values, referenced \texttt{{objects}} (which
properties are listed in its own entity), or a JSON \texttt{[list]} of these.
If processed as JSON-LD, this forms an RDF graph by matching the \texttt{@id}
IRIs and applying the \texttt{@context} mapping to \surl{schema.org} terms.
\normalsize

@@

RO-Crate reuses vocabulary terms and IRIs from \surl{Schema.org}, but provides
its own versioned JSON-LD
context,\footnote{\url{https://w3id.org/ro/crate/1.1/context}} which has a flat list
with the mapping from JSON-LD keys to their URI equivalents (e.g. key
\texttt{"author"} maps to the \url{http://schema.org/author} property).

The rationale behind this decision is to support JSON-based RO-Crate
applications that are largely unaware of JSON-LD, that still may want
to process the \texttt{@context} to find or add Linked Data definitions of
otherwise unknown properties and types. Not reusing the official
\surl{Schema.org} context means RO-Crate is also able to map in additional
vocabularies where needed, namely the \textit{Portland Common Data Model}
(PCDM) \cite{pcdm} for repositories and Bioschemas \cite{bioschemas_2017} for
describing computational workflows. RO-Crate profiles may
extend\footnote{\url{https://www.researchobject.org/ro-crate/1.1/appendix/jsonld.html\#extending-ro-crate}}
the \texttt{@context} to re-use additional domain-specific ontologies.

Similarly, while the \surl{Schema.org} context
currently\footnote{\url{https://schema.org/version/13.0/schemaorg-current-http.jsonld}}
have \texttt{"@type": "@id"} annotations for object properties, RO-Crate
JSON-LD distinguishes explicitly between references to other entities
(\texttt{{"@id": "\#alice"}})
and string values (\texttt{"Alice"}) -- meaning
RO-Crate applications can find references for corresponding entities
and IRIs without parsing the \texttt{@context} to understand a particular
property. Notably this is exploited by the \texttt{ro-crate-html-js}
\cite{ro-crate-html-js} tool to provide reliable HTML rendering for
otherwise unknown properties and types.

@@

Simplifying from the CWLProv approach, the planned Workflow Run
RO-Crate profile will use a high level \surl{Schema.org}
provenance\footnote{\url{https://www.researchobject.org/ro-crate/1.1/provenance.html\#software-used-to-create-files}}
for the input/output boundary of the overall workflow execution. This
\textit{Level 1 workflow provenance} \cite{doi:10.1093/gigascience/giz095} can be
expressed generally across workflow languages with minimal workflow
engine changes, with the option of more detailed provenance traces as
separate PROV artefacts in the RO-Crate as data entities. In the
current development of Specimen Data
Refinery\footnote{\url{https://github.com/DiSSCo/SDR}} \cite{doi:10.3897/rio.6.e57602}
these RO-Crates will document the text recognition workflow runs of
digitised biological specimens, exposed as FAIR Digital Objects
\cite{doi:10.3390/publications8020021}.

@@

The PARADISEC use case takes advantage of several RO-Crate features and
principles. Firstly, the transcribed metadata are now independent of
the PARADISEC platform and can be archived, preserved and processed in
its own right, using \surl{Schema.org} as base vocabulary and extended with
PARADISEC-specific terms.

@@

Particularly relevant to this article is the second objective of the
Harvard Data Commons, which aims to support the deposit of research
artefacts to Harvard Dataverse with sufficient information in the
metadata to allow their future reuse (Fig.~\ref{fig:hdc}). To support
the incorporation of data, code, and other artefacts from various
institutional infrastructures, Harvard Data Commons is currently
working on RO-Crate adaptation. The RO-Crate metadata provides the
necessary structure to make all research artefacts FAIR. The Dataverse
software already has extensive support for metadata, including the Data
Documentation Initiative (DDI), Dublin Core, DataCite, and \surl{Schema.org}.
Incorporating RO-Crate, which has the flexibility to describe a wide
range of research resources, will facilitate their seamless transition
from one infrastructure to the other within the Harvard Data Commons.

@@

The Digital Object Interface Protocol \cite{doip2.0} can be considered an
``abstract protocol'' of requirements, DOs could be implemented in
multiple ways. One suggested implementation is the FAIR Digital Object
Framework,\footnote{\url{https://fairdigitalobjectframework.org/}} based on HTTP and
the Linked Data Principles. While there is agreement on using PIDs
based on DOIs, consensus on how to represent common metadata, core
types and collections as FDOs has not yet been reached. We argue that
RO-Crate can play an important role for FDOs:
\begin{enumerate}
\item[1.] By providing a predictable and extensible serialisation of
structured metadata.
\item[2.] By formalising how to aggregate digital objects as collections (and
adding their context).
\item[3.] By providing a natural Metadata FDO in the form of the RO-Crate
Metadata File.
\item[4.] By being based on Linked Data and \surl{schema.org} vocabulary, meaning
that PIDs already exist for common types and properties.
\end{enumerate}

@@

RO-Crate attempts to strike a balance between these tensions, and
provides a common metadata framework that encourages extensions.
However, just like the RO-Crate specification can be thought of as a
\textit{core profile} of \surl{schema.org} in JSON-LD, we cannot stress the
importance of also establishing domain-specific RO-Crate profiles and
conventions, as explored in Sections~\ref{sec:profiles} and \ref
{sec:inuse}. Specialization comes hand-in-hand with the principle of
\textit{graceful degradation}; RO-Crate applications and users are free to
choose the semantic detail level they participate at, as long as they
follow the common syntactic requirements.

@@

The direction of future RO-Crate work is determined by the community
around it as a collaborative effort. We currently plan on further
outreach, building training material (including a comprehensive
entry-level tutorial) and maturing the reference implementation
libraries. We will also collect and build examples of RO-Crate
\textit{consumption}, e.g. Jupyter Notebooks that query multiple crates using
knowledge graphs. In addition, we are exploring ways to support some
entity types requested by users, e.g. detailed workflow runs or
container provenance, which do not have a good match in \surl{Schema.org}.
Such support could be added, for instance, by integrating other
vocabularies or by having separated (but linked) metadata files.

Furthermore, we want to better understand how the community uses
RO-Crate in practice and how it contrasts with other related efforts;
this will help us to improve our specification and tools. By
discovering commonalities in emerging usage (e.g. additional \surl{schema.org}
types), the community helps to reduce divergence that could otherwise
occur with proliferation of further RO-Crate profiles. We plan to
gather feedback via user studies, with the Linked Open Data community
or as part of EOSC Bring-your-own-Data training events.

@@

The main addition in the upcoming 1.2 release of the RO-Crate
specifications will be the formalization of
profiles\footnote{\url{https://www.researchobject.org/ro-crate/1.2-DRAFT/profiles}}
for different categories of crates. Additional entity types have been
requested by users, e.g. workflow runs, business workflows, containers
and software packages, tabular data structures; these are not always
matched well with existing \surl{schema.org} types but may benefit from other
vocabularies or even separate metadata files, e.g. from Frictionless
Data.\footnote{\url{https://frictionlessdata.io/}} We will be further aligning with
and collaborating with related research artefact description efforts
like CodeMeta\footnote{\url{https://codemeta.github.io/}} for software metadata,
Science-on-schema.org\footnote{\url{https://science-on-schema.org/}}
\cite{doi:10.5281/zenodo.4477164} for datasets, FAIR Digital
Objects\footnote{\url{https://fairdo.org/}} \cite{doi:10.3390/publications8020021} and
activities in EOSC task forces\footnote{\url{https://www.eosc.eu/task-force-faq}}
including the EOSC Interoperability Framework \cite{doi:10.2777/620649}.

@@

In reality many additional attributes and contextual types from
\surl{schema.org} types like \url{http://schema.org/affiliation} and \url
{http://schema.org/Organization} would be used to further describe the
RO-Crate and its entities, but as these are optional (\textit{SHOULD}
requirements) they do not form part of this formalization.

@@

Below follows a mapping from $\mathcal{L}_{\mathit{rocrate}}$ to $\mathcal
{L}_{\textit{RDF}}$ using \surl{schema.org} as vocabulary:
%
{\fontsize{9.6}{11.6}\selectfont{\begin{eqnarray*}
\textit{Property}(p) & \Rightarrow& \textit{type}(p, \texttt
{<http://www.w3.org/2000/01/rdf-schema\#Property>})
\\
\textit{Class}(c) & \Rightarrow& \textit{type}(c, \texttt {<http://www.w3.org/2000/01/rdf-schema
\#Class>})
\\
\textit{Dataset}(d) & \Rightarrow& \textit{type}(d, \texttt {<http://schema.org/Dataset>})
\\
\textit{File}(f) & \Rightarrow& \textit{type}(f, \texttt {<http://schema.org/MediaObject>})
\\
\textit{ContextualEntity}(e) & \Rightarrow& \textit{type}(f, \texttt
{<http://schema.org/Thing>})
\\
\textit{CreativeWork}(e) & \Rightarrow& \textit{ContextualEntity}(e) \land
\textit{type}(e, \texttt{<http://schema.org/CreativeWork>})
\\
\textit{hasPart}(e, t) & \Rightarrow& \textit{Relation}(e, \texttt
{<http://schema.org/hasPart>}, t)
\\
\textit{name}(e, n) & \Rightarrow& \textit{Attribute}(e, \texttt
{<http://schema.org/name>}, n)
\\
\textit{description}(e, s) & \Rightarrow& \textit{Attribute}(e, \texttt
{<http://schema.org/description>}, s)
\\
\textit{datePublished}(e, d) & \Rightarrow& Attribute(e, \texttt
{<http://schema.org/datePublished>}, d)
\\
\textit{license}(e, l) & \Rightarrow& \textit{Relation}(e, \texttt
{<http://schema.org/license>}, l) \land \textit{CreativeWork}(l)
\\
\textit{type}(e, t) & \Rightarrow& \textit{Relation}(e, \texttt
{<http://www.w3.org/1999/02/22-rdf-syntax-ns\#type>}, t)
\\
& & {}\land \textit{Class}(t)
\\
\textit{String}(s) & \equiv& \textit{Value}(s) \land s \in\mathbb{S}
\\
\textit{String}(s) & \Rightarrow& \textit{Datatype}(s, \texttt
{<http://www.w3.org/2001/XMLSchema\#string>})
\\
\textit{Decimal}(d) & \equiv& \textit{Value}(d) \land d \in\mathbb{R}
\\
\textit{Decimal}(d) & \Rightarrow& \textit{Datatype}(d, \texttt
{<http://www.w3.org/2001/XMLSchema\#decimal>})
\\
\textit{Relation}(s,p,o) & \Rightarrow& \textit{Triple}(s,p,o) \land
\textit{IRI}(s) \land \textit{IRI}(o)
\\
\textit{Attribute}(s,p,o) & \Rightarrow& \textit{Triple}(s,p,o) \land
\textit{IRI}(s) \land \textit{Literal}(o)
\end{eqnarray*}}}

Note that in the JSON-LD serialization of RO-Crate, the expression of
$\textit{Class}$ and $\textit{Property}$ is typically indirect: The JSON-LD \texttt
{@context} maps to \surl{schema.org} IRIs, which, when resolved as Linked
Data, embed their formal definition as RDFa. Extensions may however
include such term definitions directly in the RO-Crate.

@@

Combining the above predicates and \surl{schema.org} mapping with rudimentary
JSON templates, these forward-chaining production rules can output
JSON-LD according to the RO-Crate 1.1 specification:\footnote{\textbf{Limitations:}
 Contextual entities not related from the
RO-Crate (e.g. using inverse relations to a data entity) would not be
covered by the single direction $Mentions(R, s)$ production rule; see
\href{https://github.com/ResearchObject/ro-crate/issues/122}{GitHub
issue ResearchObject/ro-crate\#122}. The $\textit{datePublished}(e, d)$ rule do
not include syntax checks for the ISO 8601 datetime format. Compared
with RO-Crate examples, this generated JSON-LD does not use a
$\textit{@context}$ as the IRIs are produced unshortened, a post-step could do
JSON-LD Flattening with a versioned RO-Crate context. The \texttt
{@type} expansion is included for clarity, even though this is also
implied by the $\textit{type}(e, t)$ expansion to $\textit{Relation}(e, \texttt{xsd:type})$.}
%
\begin{eqnarray*}
\textit{Mentions}(R, s) \land \textit{Relation}(s,p,o) & \Rightarrow&
\textit{Mentions}(R, o)
\\
\textit{IRI}(i) & \Rightarrow& \texttt{"} i \texttt{"}
\\
\textit{Decimal}(d) & \Rightarrow& d
\\
\textit{String}(s) & \Rightarrow& \texttt{"} s \texttt{"}
\\
\forall e \forall t \ \textit{type}(e, t) & \Rightarrow& \texttt{\{"@id":} e
\texttt{,}
\\
&& \ \ \texttt{"@type":} t
\\
&& \texttt{\}}
\\
\forall s \forall p \forall o \ \textit{Relation}(s,p,o) & \Rightarrow& \texttt{
\{"@id":} s \texttt{,}
\\
&& \ \ p \texttt{: \{ "@id":} o \texttt{\}}
\\
&& \texttt{\}}
\\
\forall s \forall p \forall v \ \textit{Attribute}(s,p,v) & \Rightarrow&
\texttt{\{"@id":} s \texttt{,}
\\
&& \ \ p \texttt{:} v
\\
&& \texttt{\}}
\\
\forall r \forall c\ \textit{ROCrate}(r) & \Rightarrow& \texttt{\{
"@graph": [}
\\
&& \ \ \ \ \textit{Mentions}(r, c)*
\\
&& \ \ \ \texttt{]}
\\
&& \texttt{\}}
\\
R & \equiv& \texttt{<./>}
\\
R & \Rightarrow& \textit{MetadataFile}(\texttt{<ro-crate-metadata.json>})
\end{eqnarray*}

>>>>

\begin{review}{solicited}
\reviewer{\fnms{Angelo A.} \snm{Salatino}\ead
[label=os1,orcid]{0000-0002-4763-3943}\printead[presep={\ }]{os1}}
\reviewer{\fnms{Alessia} \snm{Bardi}\ead
[label=os2,orcid]{0000-0002-1112-1292}\printead[presep={\ }]{os2}}
\reviewer{\snm{one anonymous reviewer}}
\end{review}
%
%\dedicated{}
%
% HISTORY:
\received{\sday{13} \smonth{8} \syear{2021}}
\accepted{\sday{23} \smonth{9} \syear{2021}}
%
\begin{abstract}
%\parttitle{}
An increasing number of researchers support reproducibility by
including pointers to and
descriptions of datasets, software and methods in their publications.
However, scientific
articles may be ambiguous, incomplete and difficult to process by
automated systems. In
this paper we introduce RO-Crate, an open, community-driven, and
lightweight approach
to packaging research artefacts along with their metadata in a machine
readable manner.
RO-Crate is based on Schema.org annotations in JSON-LD, aiming to
establish best
practices to formally describe metadata in an accessible and practical
way for their
use in a wide variety of situations.

@@

For our development methodology we followed the mantra of working
closely with a small group to get a deep understanding of
requirements and ensure rapid feedback loops. We created a pool of
early adopter projects from a range of disciplines and groups,
primarily addressing developers of platforms. Thus the base level for
simplicity was \textbf{developer friendliness}.

@@

Addressing the simplicity of understanding and engaging with RO-Crate
by data practitioners is through the platforms, for example with
interactive tools (Section~\ref{sec:tooling}) like
Describo\footnote{\url{https://arkisto-platform.github.io/describo/}} \cite{describo} and
Jupyter notebooks \cite{doi:10.3233/978-1-61499-649-1-87}, and by close
discussions with domain scientists on how to appropriately capture what
they determine to be relevant metadata. This ultimately requires a new
type of awareness and learning material separate from developer
specifications, focusing on the simplicity of extensibility to serve
the user needs, along with user-driven development of new RO-Crate
Profiles specific for their needs (Section~\ref{sec:inuse}).

@@

\begin{figure}%[t!]
\includegraphics{210053f02}
\caption{Simplified UML class diagram of RO-Crate. The \emph
{RO-Crate Metadata File} conforms to a version of the specification;
and contains a JSON-LD graph \cite{sporny_2014} that describes the
entities that make up the RO-Crate. The \emph{RO-Crate Root Data
Entity} represent the Research Object as a dataset. The RO-Crate
aggregates \emph{data entities} (\texttt{hasPart}) which are further
described using \emph{contextual entities} (which may include
aggregated and non-aggregated data entities). Multiple types and
relations from Schema.org allow annotations to be more specific,
including figures, nested datasets, computational workflows, people,
organisations, instruments and places. Contextual entities not
otherwise cross-referenced from other entities' properties (\emph
{describes}) can be grouped under the root entity (\texttt{mentions}).}
\label{fig:uml}
\end{figure}

@@

As such, the RO-Crate 1.1\footnote{\url{https://w3id.org/ro/crate/1.1}}
specification \cite{doi:10.5281/zenodo.4541002} can be seen as an
opinionated and example-driven guide to writing
Schema.org\footnote{\url{https://schema.org/}} \cite{doi:10.1145/2857274.2857276}
metadata as JSON-LD \cite{sporny_2014} (see Section~\ref{sec:implementation}), which leaves it open for implementers to include
additional metadata using other Schema.org types and properties, or
even additional Linked Data vocabularies/ontologies or their own ad-hoc terms.

@@

The RO-Crate conceptual model has been realised using JSON-LD and
Schema.org in a prescriptive form as discussed in Section~\ref{sec:conceptual}. These technical choices were made to cater for
simplicity from a developer perspective (as introduced in Section~\ref{sec:methodology}).

@@

In\footnotetext{Recommended properties for types shown in
Listing \ref{lis1} also include
\texttt{affiliation}, \texttt{citation}, \texttt{contactPoint},
\texttt{description},
\texttt{encodingFormat}, \texttt{funder}, \texttt{geo}, \texttt{identifier},
\texttt{keywords},
\texttt{publisher}; these properties and corresponding contextual
entities are
excluded here for brevity. See complete example
\url
{https://www.researchobject.org/2021-packaging-research-artefacts-with-ro-crate/listing1/}.} this flattened profile of JSON-LD, each \texttt{{entity}} are directly
under \texttt{@graph} and represents the RDF triples with a common \textit{subject}
(\texttt{@id}), mapped \textit{properties} like \texttt{hasPart}, and \textit{objects} -- as
either literal \texttt{"string"} values, referenced \texttt{{objects}} (which
properties are listed in its own entity), or a JSON \texttt{[list]} of these.
If processed as JSON-LD, this forms an RDF graph by matching the \texttt{@id}
IRIs and applying the \texttt{@context} mapping to Schema.org terms.
\normalsize

@@

RO-Crate reuses vocabulary terms and IRIs from Schema.org, but provides
its own versioned JSON-LD
context,\footnote{\url{https://w3id.org/ro/crate/1.1/context}} which has a flat list
with the mapping from JSON-LD keys to their URI equivalents (e.g. key
\texttt{"author"} maps to the \url{http://schema.org/author} property).

The rationale behind this decision is to support JSON-based RO-Crate
applications that are largely unaware of JSON-LD, that still may want
to process the \texttt{@context} to find or add Linked Data definitions of
otherwise unknown properties and types. Not reusing the official
Schema.org context means RO-Crate is also able to map in additional
vocabularies where needed, namely the \textit{Portland Common Data Model}
(PCDM) \cite{pcdm} for repositories and Bioschemas \cite{bioschemas_2017} for
describing computational workflows. RO-Crate profiles may
extend\footnote{\url{https://www.researchobject.org/ro-crate/1.1/appendix/jsonld.html\#extending-ro-crate}}
the \texttt{@context} to re-use additional domain-specific ontologies.

Similarly, while the Schema.org context
currently\footnote{\url{https://schema.org/version/13.0/schemaorg-current-http.jsonld}}
have \texttt{"@type": "@id"} annotations for object properties, RO-Crate
JSON-LD distinguishes explicitly between references to other entities
(\texttt{{"@id": "\#alice"}})
and string values (\texttt{"Alice"}) -- meaning
RO-Crate applications can find references for corresponding entities
and IRIs without parsing the \texttt{@context} to understand a particular
property. Notably this is exploited by the \texttt{ro-crate-html-js}
\cite{ro-crate-html-js} tool to provide reliable HTML rendering for
otherwise unknown properties and types.

@@

Simplifying from the CWLProv approach, the planned Workflow Run
RO-Crate profile will use a high level Schema.org
provenance\footnote{\url{https://www.researchobject.org/ro-crate/1.1/provenance.html\#software-used-to-create-files}}
for the input/output boundary of the overall workflow execution. This
\textit{Level 1 workflow provenance} \cite{doi:10.1093/gigascience/giz095} can be
expressed generally across workflow languages with minimal workflow
engine changes, with the option of more detailed provenance traces as
separate PROV artefacts in the RO-Crate as data entities. In the
current development of Specimen Data
Refinery\footnote{\url{https://github.com/DiSSCo/SDR}} \cite{doi:10.3897/rio.6.e57602}
these RO-Crates will document the text recognition workflow runs of
digitised biological specimens, exposed as FAIR Digital Objects
\cite{doi:10.3390/publications8020021}.

@@

The PARADISEC use case takes advantage of several RO-Crate features and
principles. Firstly, the transcribed metadata are now independent of
the PARADISEC platform and can be archived, preserved and processed in
its own right, using Schema.org as base vocabulary and extended with
PARADISEC-specific terms.

@@

Particularly relevant to this article is the second objective of the
Harvard Data Commons, which aims to support the deposit of research
artefacts to Harvard Dataverse with sufficient information in the
metadata to allow their future reuse (Fig.~\ref{fig:hdc}). To support
the incorporation of data, code, and other artefacts from various
institutional infrastructures, Harvard Data Commons is currently
working on RO-Crate adaptation. The RO-Crate metadata provides the
necessary structure to make all research artefacts FAIR. The Dataverse
software already has extensive support for metadata, including the Data
Documentation Initiative (DDI), Dublin Core, DataCite, and Schema.org.
Incorporating RO-Crate, which has the flexibility to describe a wide
range of research resources, will facilitate their seamless transition
from one infrastructure to the other within the Harvard Data Commons.

@@

The Digital Object Interface Protocol \cite{doip2.0} can be considered an
``abstract protocol'' of requirements, DOs could be implemented in
multiple ways. One suggested implementation is the FAIR Digital Object
Framework,\footnote{\url{https://fairdigitalobjectframework.org/}} based on HTTP and
the Linked Data Principles. While there is agreement on using PIDs
based on DOIs, consensus on how to represent common metadata, core
types and collections as FDOs has not yet been reached. We argue that
RO-Crate can play an important role for FDOs:
\begin{enumerate}
\item[1.] By providing a predictable and extensible serialisation of
structured metadata.
\item[2.] By formalising how to aggregate digital objects as collections (and
adding their context).
\item[3.] By providing a natural Metadata FDO in the form of the RO-Crate
Metadata File.
\item[4.] By being based on Linked Data and the Schema.org vocabulary, meaning
that PIDs already exist for common types and properties.
\end{enumerate}

@@

RO-Crate attempts to strike a balance between these tensions, and
provides a common metadata framework that encourages extensions.
However, just like the RO-Crate specification can be thought of as a
\textit{core profile} of Schema.org in JSON-LD, we cannot stress the
importance of also establishing domain-specific RO-Crate profiles and
conventions, as explored in Sections~\ref{sec:profiles} and \ref
{sec:inuse}. Specialization comes hand-in-hand with the principle of
\textit{graceful degradation}; RO-Crate applications and users are free to
choose the semantic detail level they participate at, as long as they
follow the common syntactic requirements.

@@

The direction of future RO-Crate work is determined by the community
around it as a collaborative effort. We currently plan on further
outreach, building training material (including a comprehensive
entry-level tutorial) and maturing the reference implementation
libraries. We will also collect and build examples of RO-Crate
\textit{consumption}, e.g. Jupyter Notebooks that query multiple crates using
knowledge graphs. In addition, we are exploring ways to support some
entity types requested by users, e.g. detailed workflow runs or
container provenance, which do not have a good match in Schema.org.
Such support could be added, for instance, by integrating other
vocabularies or by having separated (but linked) metadata files.

Furthermore, we want to better understand how the community uses
RO-Crate in practice and how it contrasts with other related efforts;
this will help us to improve our specification and tools. By
discovering commonalities in emerging usage (e.g. additional Schema.org
types), the community helps to reduce divergence that could otherwise
occur with proliferation of further RO-Crate profiles. We plan to
gather feedback via user studies, with the Linked Open Data community
or as part of EOSC Bring-your-own-Data training events.

@@

The main addition in the upcoming 1.2 release of the RO-Crate
specifications will be the formalization of
profiles\footnote{\url{https://www.researchobject.org/ro-crate/1.2-DRAFT/profiles}}
for different categories of crates. Additional entity types have been
requested by users, e.g. workflow runs, business workflows, containers
and software packages, tabular data structures; these are not always
matched well with existing Schema.org types but may benefit from other
vocabularies or even separate metadata files, e.g. from Frictionless
Data.\footnote{\url{https://frictionlessdata.io/}} We will be further aligning with
and collaborating with related research artefact description efforts
like CodeMeta\footnote{\url{https://codemeta.github.io/}} for software metadata,
Science-on-schema.org\footnote{\url{https://science-on-schema.org/}}
\cite{doi:10.5281/zenodo.4477164} for datasets, FAIR Digital
Objects\footnote{\url{https://fairdo.org/}} \cite{doi:10.3390/publications8020021} and
activities in EOSC task forces\footnote{\url{https://www.eosc.eu/task-force-faq}}
including the EOSC Interoperability Framework \cite{doi:10.2777/620649}.

@@

In reality many additional attributes and contextual types from
Schema.org types like \url{http://schema.org/affiliation} and \url
{http://schema.org/Organization} would be used to further describe the
RO-Crate and its entities, but as these are optional (\textit{SHOULD}
requirements) they do not form part of this formalization.

@@

Below follows a mapping from $\mathcal{L}_{\mathit{rocrate}}$ to $\mathcal
{L}_{\textit{RDF}}$ using Schema.org as vocabulary:
%
{\fontsize{9.6}{11.6}\selectfont{\begin{eqnarray*}
\textit{Property}(p) & \Rightarrow& \textit{type}(p, \texttt
{<http://www.w3.org/2000/01/rdf-schema\#Property>})
\\
\textit{Class}(c) & \Rightarrow& \textit{type}(c, \texttt {<http://www.w3.org/2000/01/rdf-schema
\#Class>})
\\
\textit{Dataset}(d) & \Rightarrow& \textit{type}(d, \texttt {<http://schema.org/Dataset>})
\\
\textit{File}(f) & \Rightarrow& \textit{type}(f, \texttt {<http://schema.org/MediaObject>})
\\
\textit{ContextualEntity}(e) & \Rightarrow& \textit{type}(f, \texttt
{<http://schema.org/Thing>})
\\
\textit{CreativeWork}(e) & \Rightarrow& \textit{ContextualEntity}(e) \land
\textit{type}(e, \texttt{<http://schema.org/CreativeWork>})
\\
\textit{hasPart}(e, t) & \Rightarrow& \textit{Relation}(e, \texttt
{<http://schema.org/hasPart>}, t)
\\
\textit{name}(e, n) & \Rightarrow& \textit{Attribute}(e, \texttt
{<http://schema.org/name>}, n)
\\
\textit{description}(e, s) & \Rightarrow& \textit{Attribute}(e, \texttt
{<http://schema.org/description>}, s)
\\
\textit{datePublished}(e, d) & \Rightarrow& Attribute(e, \texttt
{<http://schema.org/datePublished>}, d)
\\
\textit{license}(e, l) & \Rightarrow& \textit{Relation}(e, \texttt
{<http://schema.org/license>}, l) \land \textit{CreativeWork}(l)
\\
\textit{type}(e, t) & \Rightarrow& \textit{Relation}(e, \texttt
{<http://www.w3.org/1999/02/22-rdf-syntax-ns\#type>}, t)
\\
& & {}\land \textit{Class}(t)
\\
\textit{String}(s) & \equiv& \textit{Value}(s) \land s \in\mathbb{S}
\\
\textit{String}(s) & \Rightarrow& \textit{Datatype}(s, \texttt
{<http://www.w3.org/2001/XMLSchema\#string>})
\\
\textit{Decimal}(d) & \equiv& \textit{Value}(d) \land d \in\mathbb{R}
\\
\textit{Decimal}(d) & \Rightarrow& \textit{Datatype}(d, \texttt
{<http://www.w3.org/2001/XMLSchema\#decimal>})
\\
\textit{Relation}(s,p,o) & \Rightarrow& \textit{Triple}(s,p,o) \land
\textit{IRI}(s) \land \textit{IRI}(o)
\\
\textit{Attribute}(s,p,o) & \Rightarrow& \textit{Triple}(s,p,o) \land
\textit{IRI}(s) \land \textit{Literal}(o)
\end{eqnarray*}}}

Note that in the JSON-LD serialization of RO-Crate, the expression of
$\textit{Class}$ and $\textit{Property}$ is typically indirect: The JSON-LD \texttt
{@context} maps to Schema.org IRIs, which, when resolved as Linked
Data, embed their formal definition as RDFa. Extensions may however
include such term definitions directly in the RO-Crate.

@@

Combining the above predicates and Schema.org mapping with rudimentary
JSON templates, these forward-chaining production rules can output
JSON-LD according to the RO-Crate 1.1 specification:\footnote{\textbf{Limitations:}
 Contextual entities not related from the
RO-Crate (e.g. using inverse relations to a data entity) would not be
covered by the single direction $Mentions(R, s)$ production rule; see
\href{https://github.com/ResearchObject/ro-crate/issues/122}{GitHub
issue ResearchObject/ro-crate\#122}. The $\textit{datePublished}(e, d)$ rule do
not include syntax checks for the ISO 8601 datetime format. Compared
with RO-Crate examples, this generated JSON-LD does not use a
$\textit{@context}$ as the IRIs are produced unshortened, a post-step could do
JSON-LD Flattening with a versioned RO-Crate context. The \texttt
{@type} expansion is included for clarity, even though this is also
implied by the $\textit{type}(e, t)$ expansion to $\textit{Relation}(e, \texttt{xsd:type})$.}
%
\begin{eqnarray*}
\textit{Mentions}(R, s) \land \textit{Relation}(s,p,o) & \Rightarrow&
\textit{Mentions}(R, o)
\\
\textit{IRI}(i) & \Rightarrow& \texttt{"} i \texttt{"}
\\
\textit{Decimal}(d) & \Rightarrow& d
\\
\textit{String}(s) & \Rightarrow& \texttt{"} s \texttt{"}
\\
\forall e \forall t \ \textit{type}(e, t) & \Rightarrow& \texttt{\{"@id":} e
\texttt{,}
\\
&& \ \ \texttt{"@type":} t
\\
&& \texttt{\}}
\\
\forall s \forall p \forall o \ \textit{Relation}(s,p,o) & \Rightarrow& \texttt{
\{"@id":} s \texttt{,}
\\
&& \ \ p \texttt{: \{ "@id":} o \texttt{\}}
\\
&& \texttt{\}}
\\
\forall s \forall p \forall v \ \textit{Attribute}(s,p,v) & \Rightarrow&
\texttt{\{"@id":} s \texttt{,}
\\
&& \ \ p \texttt{:} v
\\
&& \texttt{\}}
\\
\forall r \forall c\ \textit{ROCrate}(r) & \Rightarrow& \texttt{\{
"@graph": [}
\\
&& \ \ \ \ \textit{Mentions}(r, c)*
\\
&& \ \ \ \texttt{]}
\\
&& \texttt{\}}
\\
R & \equiv& \texttt{<./>}
\\
R & \Rightarrow& \textit{MetadataFile}(\texttt{<ro-crate-metadata.json>})
\end{eqnarray*}
